---
title: "HW4"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = "~/Desktop/CIS 576/HW/HW 4 and Data Set")
```

That tells the knitter to look for files in that exact folder. Then try knitting again!
```

## Setup

Install and load packages.

```{r libraries}
#install.packages(c("tidyverse", "tidygraph", "ggraph", "readxl", "treemapify"))

library(tidyverse)
library(tidygraph)
library(ggraph)
library(readxl)
library(treemapify)
```

---

# Part 1: Network Visualization — Employee Email Communication

## Steps Taken

The email dataset contains two sheets: `Departments` (30 employees across 6 departments: Marketing, HR, Finance, Operations, Sales, and Purchase) and `Emails` (149 directed edges with a frequency variable indicating how many emails were exchanged). I loaded both sheets and combined them into a `tbl_graph` object using `tidygraph`, treating the data as a directed network. I then used `ggraph` to create three meaningfully different visualizations. The first uses a standard node-link layout to reveal the overall network structure and department clustering. The second encodes email frequency via edge width and arrow direction to highlight communication intensity. The third uses a linear arc diagram to reduce visual clutter and make individual communication patterns easier to trace.

```{r network-data}
# Load data
edges <- read_excel("EmployeeEmails.xlsx", sheet = "Emails")
nodes <- read_excel("EmployeeEmails.xlsx", sheet = "Departments")

# Rename for clarity
nodes <- nodes %>% rename(name = employee)

# Build directed network
network <- tbl_graph(nodes = nodes, edges = edges, directed = TRUE)
network
```

---

### Network Visualization 1:  Node-Link Diagram

**Description:**displays all 30 employees as nodes and 149 emails. The nodes are colored by department making it easy to spot which departments are most connected across departments. 

```{r network-viz1, fig.width=9, fig.height=7}
ggraph(network, layout = "nicely") +
  geom_edge_link(color = "gray75", alpha = 0.5,
                 arrow = arrow(length = unit(2.5, "mm"), type = "closed"),
                 end_cap = circle(3, "mm")) +
  geom_node_point(aes(color = department), size = 5) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_color_brewer(palette = "Set2") +
  theme_graph(base_family = "sans") +
  labs(title = "Employee Email Communication Network",
       subtitle = "Nodes colored by department; arrows indicate email direction",
       color = "Department")
```

---

### Network Visualization 2: Edge Width Scaled by Email Frequency

**Description:** Email frequency as a visual encoding. Edge width and transparency are both by the number of emails exchanged between two employees. The thicker and darker edges indicate higher communication frequency. This shows which employees communicate most often by showing more than just who is connected to whom. This satisfies the requirement of incorporating email frequency.

```{r network-viz2, fig.width=9, fig.height=7}
ggraph(network, layout = "nicely") +
  geom_edge_link(aes(width = frequency, alpha = frequency),
                 color = "steelblue",
                 arrow = arrow(length = unit(3, "mm"), type = "closed"),
                 end_cap = circle(3, "mm")) +
  scale_edge_width(range = c(0.2, 2.5), name = "Email Frequency") +
  scale_edge_alpha(range = c(0.1, 0.85), guide = "none") +
  geom_node_point(aes(color = department), size = 5) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_color_brewer(palette = "Set2") +
  theme_graph(base_family = "sans") +
  labs(title = "Email Frequency Between Employees",
       subtitle = "Edge width and opacity scaled by number of emails sent",
       color = "Department")
```

---

### Network Visualization 3: Arc Diagram

**Description:** This arc arranges all employees along a curved arc to represent email exchanges, with arc thickness scaled by frequency. Employees are colored by department, and the arc height shows directionality. This is a meaningfully different representation from the node-link diagrams above.

```{r network-viz3, fig.width=11, fig.height=6}
ggraph(network, layout = "linear") +
  geom_edge_arc(aes(width = frequency, alpha = frequency),
                color = "steelblue", fold = TRUE) +
  scale_edge_width(range = c(0.2, 2), name = "Email Frequency") +
  scale_edge_alpha(range = c(0.1, 0.8), guide = "none") +
  geom_node_point(aes(color = department), size = 4) +
  geom_node_text(aes(label = name), angle = 65, hjust = 1,
                 vjust = 1, size = 2.8) +
  scale_color_brewer(palette = "Set2") +
  theme_graph(base_family = "sans") +
  labs(title = "Arc Diagram of Employee Email Communication",
       subtitle = "Arc thickness scaled by email frequency; employees arranged linearly",
       color = "Department")
```

---

# Part 2: Hierarchical Visualization — Magnoliopsida Plants in Colorado

## Steps Taken

The Colorado plants dataset contains 1,787 plant records with a Class, Order, Family, Genus hierarchy (34 Orders, 72 Families, 370 Genera). I excluded the Scientific Name level to keep the visualizations readable. I first deduplicated the data to Class–Order–Family–Genus combinations, then built a list by combining the three levels of parent–child relationships into a single graph. Three visualizations — vertical tree, circular dendrogram, and treemap.

```{r hierarchical-data}
# Load and prepare data
plants <- read_csv("COPlants_Magnoliopsida.csv") %>%
  select(Class, Order, Family, Genus) %>%
  distinct()

# Build edge list for each hierarchy level
edges_co <- bind_rows(
  plants %>% distinct(from = Class,  to = Order),
  plants %>% distinct(from = Order,  to = Family),
  plants %>% distinct(from = Family, to = Genus)
)

# Build node list
nodes_co <- tibble(name = unique(c(edges_co$from, edges_co$to)))

# Create tbl_graph
hier_network <- tbl_graph(nodes = nodes_co, edges = edges_co, directed = TRUE)
hier_network
```

---

### Hierarchical Visualization 1: Vertical Tree Diagram

**Description:** The tree diagram displays Class, Order, Family, and Genus hierarchy in a top down layout. The tree makes it easy to see how many families each order contains and how the hierarchy fans out from the single class at the root. Leaf nodes (genera) are colored green to distinguish them from internal nodes.

```{r hier-viz1, fig.width=12, fig.height=9}
hier_network <- hier_network %>%
  activate(nodes) %>%
  mutate(leaf = node_is_leaf())

ggraph(hier_network, layout = "tree") +
  geom_edge_link(color = "gray70", alpha = 0.6) +
  geom_node_point(aes(color = leaf), size = 1.8) +
  scale_color_manual(values = c("TRUE" = "forestgreen", "FALSE" = "sienna"),
                     labels = c("TRUE" = "Genus (leaf)", "FALSE" = "Class/Order/Family"),
                     name = "Node Type") +
  geom_node_text(aes(filter = !leaf, label = name),
                 size = 2.5, repel = TRUE, color = "black") +
  theme_graph(base_family = "sans") +
  labs(title = "Magnoliopsida Plant Hierarchy — Vertical Tree",
       subtitle = "Class → Order → Family → Genus | Brown = internal nodes, Green = genera")
```

---

### Hierarchical Visualization 2: Circular Dendrogram

**Description:** The dendrogram wraps into a radial layout, which uses space more efficiently and gives a visually distinct perspective compared to the vertical tree. Internal nodes (Class, Order, Family) are labeled, while leaf nodes (Genera) are shown as small green points at the outer ring. 

```{r hier-viz2, fig.width=11, fig.height=11}
ggraph(hier_network, layout = "dendrogram", circular = TRUE) +
  geom_edge_diagonal(color = "gray65", alpha = 0.5) +
  geom_node_point(aes(filter = leaf), size = 1.2, color = "forestgreen") +
  geom_node_text(aes(filter = !leaf, label = name),
                 size = 2.2, repel = TRUE, color = "sienna4") +
  theme_graph(base_family = "sans") +
  labs(title = "Magnoliopsida Plant Hierarchy — Circular Dendrogram",
       subtitle = "Radial layout; green points = genera, brown labels = Class/Order/Family")
```

---

### Hierarchical Visualization 3: Treemap

**Description:** This treemap uses a =rectangular layout to show the relative size of each Family within its Order and is measured by the number of distinct genera it contains. Each rectangle represents a family, its area shows a genus count, and color shows Order. 

```{r hier-viz3, fig.width=12, fig.height=8}
treemap_data <- plants %>%
  group_by(Order, Family) %>%
  summarise(n_genera = n_distinct(Genus), .groups = "drop")

ggplot(treemap_data, aes(area = n_genera,
                          fill = Order,
                          label = paste0(Family, "\n(", n_genera, ")"),
                          subgroup = Order)) +
  geom_treemap() +
  geom_treemap_subgroup_border(color = "white", size = 2) +
  geom_treemap_subgroup_text(place = "topleft", color = "white",
                              fontface = "bold", size = 9, alpha = 0.8) +
  geom_treemap_text(color = "white", place = "centre",
                    size = 7, reflow = TRUE) +
  theme(legend.position = "none") +
  labs(title = "Treemap of Magnoliopsida Families by Order",
       subtitle = "Rectangle area = number of genera in that family | Grouped and colored by Order")
```
